---
title: "Data Science and the use of R Case Study 2"
date: "4-8-2017"
output:
  html_document:
    keep_md: yes
authors: Ivelin Angelov, Laura Bishop, Ethan Graham, Scott Gozdzialski
---


## Introduction
As we are learning about Data Science we are building a tool box to help us in pour careers.  Each of these tools have differenct uses.  Some are for analysis of the data, ANOVA, t-test, and linear regression to name a few.  Some are to visualize the data and its characteristics, Scatterplots, box plot, and histograms to name a few.  Some are to get to and clean the data, Python, r, and SAS are a few of these tools.  This study will be focusing on the R programming language and what is can do for us in our careers as data scientists.


## Question 2 of the case study
For our first example we will be using a built in data set called "Orange." Using the built in data set allows us to go straight to the analysis part of the problem.

The first question we are going to answer is the mean and medians for the orange trees of different sizes.
```{r}
# create a new data set with all the samme information as Orange
data1 <- Orange  

#Reorder the Tree size factor to 1, 2, 3, 4, 5
data1$Tree <- factor(data1$Tree, levels(data1$Tree)[c(2,4,1,5,3)])

#Create a function to print just the mean and median of the circumference
myFun <- function(x) {
  c(mean = mean(x), median = median(x))
}

#run the function and format it for a pretty display
cbind(Tree_Size = unique(data1$Tree),do.call(rbind, tapply(data1$circumference, data1$Tree, myFun)))
```

Next we are going to show you some of the ways you can plot data.  We are first going to create a scatterplot of the tree cirumferance against its age.

```{r}
plot(data1$circumference, data1$age, xlab = "Circumference", ylab = "Age")
print("\n") #used for formatting of the line.
```

And the next plot is a boxplot or whiskerplot, but to make things more difficult programming wise we are going to order the plots in increasing order of maximum diameter. 
```{r}
#Reorder the Tree size factor to 3, 1, 5, 2, 4
data1$Tree <- factor(data1$Tree, levels(data1$Tree)[c(3,1,5,2,4)])

#plot the temp data set circumference vs Tree
boxplot(circumference ~ Tree, data = data1)
```