---
title: "Data Science and the use of R Case Study 2"
date: "4-8-2017"
output:
  html_document:
    keep_md: yes
authors: Ivelin Angelov, Laura Bishop, Ethan Graham, Scott Gozdzialski
---


## Introduction
As we are learning about Data Science we are building a tool box to help us in our careers.  Each of these tools have differenct uses.  Some are for analysis of the data, like ANOVA, t-test, and linear regression.  Others help with visualizing the data and its characteristics, like scatterplots, box plot, and histograms.  Some tools focus on retrieving and cleaning the data, like Python, R, and SAS.  This study will be focusing on the R programming language and what value R brings to the role of Data Scientist.


## Question 2 of the case study
For our first example we will be using a built in data set called "Orange." Using the built in data set allows us to go straight to the analysis part of the problem.

The first question we are going to answer is the mean and medians for the orange trees of different sizes.
```{r}
# create a new data set with all the samme information as Orange
data1 <- Orange  

#Reorder the Tree size factor to 1, 2, 3, 4, 5
data1$Tree <- factor(data1$Tree, levels(data1$Tree)[c(2,4,1,5,3)])

#Create a function to print just the mean and median of the circumference
myFun <- function(x) {
  c(mean = mean(x), median = median(x))
}

#run the function and format it for a pretty display
cbind(Tree_Size = unique(data1$Tree),do.call(rbind, tapply(data1$circumference, data1$Tree, myFun)))
```

A scatterplot is created to show the tree circumferance as compared to age.

```{r}
plot(data1$circumference, data1$age, xlab = "Circumference", ylab = "Age")
print("\n") #used for formatting of the line.
```

The next plot is a boxplot or whiskerplot. By ordering the plots in increasing order of maximum diameter, this makes the plot easier to read but harder to create.  
```{r}
#Reorder the Tree size factor to 3, 1, 5, 2, 4
data1$Tree <- factor(data1$Tree, levels(data1$Tree)[c(3,1,5,2,4)])

#plot the temp data set circumference vs Tree
boxplot(circumference ~ Tree, data = data1)
```

Question 3 focuses on manipulating temperature data.  TEMP.csv and CityTemp.csv are very big data sets.  After cleaning, TEMP has 541,645 observations and CityTemp has 226,398.  

Question 3 - (i)  ...skipping for now.

In looking at US land temperatures, a data frame for UStemp was created along with an additional column for average land temperatures.
```{r}
UStemp <- c(subset(x=tempClean, tempClean$Date >= "1990-01-01"))
UStemp$Monthly.AverageTemp.Land <- as.numeric()
```

