---
title: "Makefile"
output: 
  html_document:
    keep_md: true
authors: Ivelin Angelov, Laura Bishop, Ethan Graham, Scott Gozdzialski
---
# This is the project's Makefile.
This file will execute the tasks: dependencies, download, clean, merge and render
All tasks are in separate cells and represent steps towards the final data analysis of the "Case Study 1" prject.
In order to keep the project modular and easy to maintain, we keep the code for all tasks into separate R files.

#### <font color='red'>In order to Knit this file you need to have R Working Directory set to the root of the project.</font>

## Dependencies
Before we start, we make sure all packages needed for the project are installed.
```{r dependencies, message=F}
needed_packages = c('downloader', 'ggplot2', 'scales', 'Hmisc', 'lattice', 'htmlTable')
for (needed_package in needed_packages) {
  if(needed_package %in% rownames(installed.packages()) == FALSE) {
    install.packages(needed_package, repos = structure(c(CRAN = "http://cran.r-project.org")))
  }
}
```

## Download
This taks is used for downloading of all necessary for the project csv data files.
The files are downloaded from their original locations as given by the assignment.
After execution two files will be created: data/gdp.csv and data/income_groups.csv
```{r download, message=F}
source(file.path("source", "download.R"), verbose=FALSE)
# For details see source/download.R
```
